---
title: "Comparing the Effectiveness of VR-Based and CBT Interventions for Autism Spectrum
  Disorder"
author: "Brendan Flaherty (D00285274) & Jeremiah Spillane (D00295490)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    df_print: paged
  html_notebook:
    toc: true
---

# Data Analytics for Immersive Environments (COMPC9047)

## Investigating the Impact of Therapeutic Interventions on Austistic Traits using the Autism Spectrum Quotient (AQ)

**Participants:** Brendan Flaherty (D00285274) & Jeremiah Spillane (D00295490) **Module:** Data Analytics for Immersive Environments (COMPC9047) **Date:** `r Sys.Date()`

------------------------------------------------------------------------

# Abstract 
## Aim and rationale 
## Participants and setting 
## Experiment design 
## Results gathering
## Findings/implications

# Introduction

## Topic and Context

Briefly describe the Autism Spectrum Disorder (ASD) as a neurodevelopmental spectrum condition and introduce the concept of neurodiversity.

## Rationale

Why are we doing this study? This study investigates which method yileds greater improvements in AQ

## Hypotesis

#### **H1:** After controlling for baseline AQ, there will be no group differences in post-treatment (null-hypothesis)

#### **H2:** Both CBT abd VR therapy groups will show significant reductions in AQ scores from pre-treatment to post-treatment, while the control group will not

#### **H3:** ???

# Method

#### **Participants:** 300 adults (M = 150, F = 150), aged 12-40. Random assignment to CBT (n = 100), VR (n = 100), Control (n = 100).

#### **Design:** Mixed 3 (Group: CBT, VR, Control - between subjects) x 2 (Time: Pre, Post - within-subjects) design.

#### **Materials:** Autism Spectrum Quotient (AQ) 50-item questionnaire, scored 0-50.1.

#### **Procedure:** Each participant completed AQ before and after 12 weeks of therapy.

-   CBT: Weekly sessions on social/emotional regulation.

-   VR: Immersive simulations for social interaction.

-   Control: No intervention.

Data recorded in columns P1_Pre_Score and P2_Post_Score.

# Results

### \*\*Data Cleaning\*

```{r Data Cleaning, message=TRUE, warning=TRUE, paged.print=FALSE}
# Adding librarys to notebook to run the code correctly
library(tidyverse)
library(ggpubr)
library(rstatix)
library(knitr)
# Read the data frames of the .csv file
df <- read_csv("CA1_Autism_VR_Study_Data_Final.csv")
# Check the column names
colnames(df)
# Change column names to lowercase
df <- df %>% rename_with(tolower) %>% mutate(group = str_trim(tolower(group)))
# Inspect Structure
glimpse(df)
summary(select(df, p1_pre_score, p2_post_score))
# Check for missing values
sapply(df, function(x) sum(is.na(x)))
# Check for values outside the scope of the questionnaire - 0 - 50.1
filter(df, p1_pre_score < 0 | p1_pre_score > 50.1 | p2_post_score < 0 | p2_post_score > 50.1)
# Drop incomplete rows
df <- df %>% drop_na(p1_pre_score, p2_post_score)
# Identify change_score (pre-post AQ)
df <- df %>% mutate(change_score = p1_pre_score - p2_post_score)
# Confirm groups
unique(df$group)
# Check for summary of all columns
summary(select(df, p1_pre_score, p2_post_score, change_score))
```

### **Descriptive Statistics**

#### **Plots**

```{r box plot, message=TRUE, warning=TRUE}
# BoxPlot created to show pre-treatment AQ scores
pre_box <- ggplot(df, aes(x = group, y = p1_pre_score, fill = group)) + geom_boxplot() + scale_fill_manual(values = c("skyblue", "lightgreen", "orange")) + labs(title = "Pre-Treatment AQ Scores",  x = "Group", y = "AQ Score (0–50)") + theme_minimal(base_size = 13) + theme(plot.title = element_text(hjust = 0.5, face = "bold"))

# BoxPlot created to show post-treatment AQ scores
post_box <- ggplot(df, aes(x = group, y = p2_post_score, fill = group)) + geom_boxplot() + scale_fill_manual(values = c("skyblue", "lightgreen", "orange")) + labs(title = "Post-Treatment AQ Scores",  x = "Group", y = "AQ Score (0–50)") + theme_minimal(base_size = 13) + theme(plot.title = element_text(hjust = 0.5, face = "bold"))

# Use ggpbr library to arrange histograms side-by-side to show difference pre and post treatment
library(ggpubr)
ggarrange(pre_box, post_box, ncol = 2, common.legend = TRUE, legend = "bottom")

# Pre-treatment AQ scores shows similar median. Post-Treatment AQ scores show VR Therapy group noticeably lower indicating greater improvement. Boxes show IQR and whiskers show outliers

```

```{r bar chart, message=TRUE, warning=TRUE}
# Summarise and group mean score data by group for creating bar chart
avg_dep_scores <- df %>% summarise(mean_pre = mean(p1_pre_score), mean_post = mean(p2_post_score), .by = group) %>% pivot_longer(cols = c(mean_pre, mean_post), names_to = "time", values_to = "mean_score") %>% mutate(time = recode(time, "mean_pre" = "Pre-Treatment", "mean_post" = "Post-Treatment"))

# Bar chart created to show mean AQ scores before and after treatment for each group
ggplot(avg_dep_scores, aes(x = group, y = mean_score, fill = time)) + geom_bar(stat = "identity", position = position_dodge()) + scale_fill_manual(values = c("orange", "skyblue")) + labs(title = "Average AQ (Depression) Scores by Group", x = "Therapy Group", y = "Mean AQ Score (0-50)" ) + theme_minimal(base_size = 13)

# Both CBT and VR Therapy groups show reduced mean AQ scores after treatment
```

```{r line-graph, message=TRUE, warning=TRUE}
# Summarise and group mean score data by time for creating line graph
avg_dep_time <- df %>% summarise(mean_pre = mean(p1_pre_score), mean_post = mean(p2_post_score), .by = group) %>% pivot_longer(cols = c(mean_pre, mean_post), names_to = "time", values_to = "mean_score") %>% mutate(time = recode(time, "mean_pre" = "Pre-Treatment", "mean_post" = "Post-Treatment"))

# Line graph showing mean AQ scores for each group pre and post treatment
ggplot(avg_dep_time, aes(x = time, y = mean_score, group = group, color = group)) + geom_line(linewidth = 1.2) + geom_point(size = 13) + scale_colour_manual(values = c("orange", "skyblue", "lightgreen")) + labs(title = "Change in Mean AQ Scores Over Time by Group", x = "Time Point", y = "Mean AQ Score (0-50)" ) + theme_minimal(base_size = 13)

# Both CBT and VR Therapy groups downward trends from pre-treatment to post-treatment - VR Therapy shows greater reduction 
```

```{r histograms, message=TRUE, warning=TRUE}
# Calculate pre-treatment AQ score distributions across the 3 groups
pre_hist <- ggplot(df, aes(x = p1_pre_score, fill = group)) + geom_histogram(binwidth = 2, alpha = 0.6, position = "identity", color = "black") + geom_vline(data = df %>% group_by(group) %>% summarise(mean_pre = mean(p1_pre_score)), aes(xintercept = mean_pre, colour = group), linetype = "dashed", linewidth = 1) +scale_fill_manual(values = c("orange", "skyblue", "lightgreen")) + scale_colour_manual(values = c("orange", "skyblue", "lightgreen")) + labs(title = "Pre-Treatment AQ Score Distribution", x = "AQ Score (0 -50)", y = "Frequency") + theme_minimal(base_size = 13) + theme(plot.title = element_text(hjust = 0.5, face = "bold"))

# Calculate post-treatment AQ score distributions across the 3 groups
post_hist <- ggplot(df, aes(x = p2_post_score, fill = group)) + geom_histogram(binwidth = 2, alpha = 0.7, colour = "black") + geom_vline(data = df %>% group_by(group) %>% summarise(mean_post = mean(p2_post_score)), aes(xintercept = mean_post, color = group), linetype = "dashed", linewidth = 1) + scale_fill_manual(values = c("orange", "skyblue", "lightgreen")) + scale_colour_manual(values = c("orange", "skyblue", "lightgreen")) + labs(title = "Post-Treatment AQ Score Distribution", x = "AQ Score (0-50)", y = "Frequency") + theme_minimal(base_size = 13) + theme(plot.title = element_text(hjust = 0.5, face = "bold"))

# Use ggpbr library to arrange histograms side-by-side to show difference pre and post treatment
library(ggpubr)
ggarrange(pre_hist, post_hist, ncol = 2, common.legend = TRUE, legend = "bottom")

# All groups show similar AQ score pre-treatment. Post-treatment, both CBT and VR Therapy groups show lower AQ score indicating improvement
```

```{r scatterplot, message=TRUE, warning=TRUE}

# Scatterplot showing relationship between pre-treatment AQ scores and change in AQ scores post-treatment
ggplot(df, aes(x = p1_pre_score, y = change_score, colour = group)) + geom_point(size = 3, alpha = 0.7) + geom_smooth(method = "lm", se = FALSE, linetype = "dashed") + scale_colour_manual(values = c("orange", "skyblue", "lightgreen")) + labs(title = "Pre-Treatment AQ vs Change in Score", x = "Pre-Treatment AQ Score", y = "Change in AQ (Pre - Post, Positive = Improvement)" ) + theme_minimal()

# A positive change indicates improvement - Participants with higher pre-treatment AQ scores tended to have a larger improvement

```

### **Inferential Statistics**

```{r paired t-test, message=TRUE, warning=TRUE}
# Assign and filter by group
cbt <- df %>% filter(group == "cbt")
vr <- df %>% filter(group == "vr")
control <- df %>% filter (group == "control")

# t-test by group
t_cbt <- t.test(cbt$p1_pre_score, cbt$p2_post_score, paired = TRUE)
t_vr <- t.test(vr$p1_pre_score, vr$p2_post_score, paired = TRUE)
t_control <- t.test(control$p1_pre_score, control$p2_post_score, paired = TRUE)

# Output results of t-tests in list
list(CBT = t_cbt, VR = t_vr, CONTROL = t_control)

# Interpretation: 
# CBT - significant improvement: t=22.9, p<.001. Mean diff = 10.99. CI between 9.2-11
# VR - Very big improvement: t=37.6, p<.001. Mean diff = 18.62. CI between 17.6-19.6
# CONTROL - no significant change: t=1.59, p=0.115. Mean diff = 0.6. CI include 0 - no change

# Results suggest that both CBT and VR reduced AQ scores with VR Therapy having a significant impact
```

```{r Between group t-test, message=TRUE, warning=TRUE}
# Assign and filter by groups - VR & Control, CBT & Control, VR and CBT
vr_control <- df %>% filter(group %in% c("vr", "control"))
cbt_control <- df %>% filter(group %in% c("cbt", "control"))
vr_cbt <- df %>% filter(group %in% c("vr", "cbt"))

# t-test between groups - VR vs Control | CBT vs Control | VR vs CBT 
t_vr_control <- t.test(change_score ~ group, data = vr_control, var.equal = TRUE)
t_cbt_control <- t.test(change_score ~ group, data = cbt_control, var.equal = TRUE)
t_vr_cbt <- t.test(change_score ~ group, data = vr_cbt, var.equal = TRUE)

# Output results of between group t-tests in list
list(VR_vs_CONTROL = t_vr_control, CBT_vs_CONTROL = t_cbt_control, VR_vs_CBT = t_vr_cbt)

# Interpretation: 
# VR vs CONTROL - VR improves more than control: t=-28.95, p<.001. Mean improvement: VR = 18.62, Control = 0.6. CI between -19.25, -16.79. 18 point improvement 18.62 - 0.6
# CBT vs CONTROL - CBT improves more than control: t=16.35, p<.001. Mean improvement: CBT = 10.09, Control = 0.6. CI between 8.35, 10.63. 9/10 point improvement 10.09 - 0.6
# VR vs CBT - CBT improves more than control: t=-12.38, p<.001. Mean improvement: VR = 18.62, CBT = 10.09. CI between -9.84, -7.22. 7-10 point improvement 18.62 - 10.09

# Results suggest that both CBT and VR were a significant improvement on the Control group and comparing VR and CBT showed that the VR group achieved a larger improvement

```

```{r z-test, message=TRUE, warning=TRUE}
# Calculate the sample mean of change score
mean_change <- mean(df$change_score)
# Calculate the standard deviation of change score
sd_change <- sd(df$change_score)
# Calculate the number of participants
n <- nrow(df)
# Calculate Z-value using the formuala
z_value <- (mean_change - 0) / (sd_change / sqrt(n))
# Calculate p-value using the Z distribution
p_value <- 2 * (1 - pnorm(abs(z_value)))
# Calculate the Confidence Interval around the mean
error_margin <- qnorm(0.975) * (sd_change / sqrt(n))
# Calculate the Lower limits of the CI
ci_lower <- mean_change - error_margin
# Calculate the Upper limits of the CI
ci_upper <- mean_change + error_margin
# Create a table for the results
data.frame(Mean_Change = round(mean_change, 2), Z_Value = round(z_value, 2), P_Value = round(p_value, 4), CI_Lower = round(ci_lower, 2), CI_Upper = round(ci_upper, 2))

# Interpretation: 
# The average change in AQ scores across all participants showed a positive value (9.77) meaning that scores decreased after treatment and that participants improved

# The Z score of 19.72 shows that the well above 0 and indicates that improvement is highly unlikely random

# P value is 0, so can reject null hypothesis

# 95% Confidence Interval of lower as 8.8 and upper limits at 10.74 shows that the true average improvement is narrow and shows consistent treatment effect

```


#### **Probability**

#### **Hypothesis Testing**

#### **Regresssion Analysis**

#### **Confidence Interval**

```{r}

```

### **Statistical tests**

### **Hypothesis 1 -**

### **Hypothesis 2 -**

### **Hypothesis 3 -**

### **Magnitude and direction of results**

# Discussion

**This is from the brief**
"The Autism Spectrum Quotient (AQ) is a 50-item self report questionnaire with a total score between 0 and 50.1. A higher score indicates a greater number of autistic traits. A total score os 26 or higher is often used as a threshold to suggest the presence of clinically significant autistic traits. Research indicates that the majority of individuals with the official autism diagnosis score 32 or above. In contrast, the average score for neurotypical males is around 17 and for neurotypical females is about 15.6 with males tending to score higher than females in the general population"

### Outline findings and relation to the hypothesis

**Paired t-test results**
Undertaken to analyse changes in AQ scores pre and post-treatment in each group
Question:
  Is there a different in each group pre and post treatment?
  
 - Null hypothesis - No significant difference (pre and post-treatment)
 - Alternative hypothesis - Significant difference (pre and post-treatment)

CBT - significant improvement: t=22.9, p<.001. Mean diff = 10.99. CI between 9.2-11
VR - Very big improvement: t=37.6, p<.001. Mean diff = 18.62. CI between 17.6-19.6
CONTROL - no significant change: t=1.59, p=0.115. Mean diff = 0.6. CI include 0 - no change

Results suggest that both CBT and VR reduced AQ scores with VR Therapy having a significant impact

**Between group t-test results**
Undertaken to analyse if there are significant differences between AQ score between CBT and VR Therapy group
Questions:
- Is VR Therapy better than no treatment (VR vs CONTROL)?
- Is CBT better than no treatment (CBT vs CONTROL)?
- Is VR Therapy more effective than CBT (VR vs CBT)?

 - Null hypothesis - No significant difference in mean improvement
 - Alternative hypothesis - One group differs in mean improvement
 
VR vs CONTROL - VR improves more than control: t=-28.95, p<.001. Mean improvement: VR = 18.62, Control = 0.6. CI between -19.25, -16.79. 18 point improvement 18.62 - 0.6 - Significant improvement

CBT vs CONTROL - CBT improves more than control: t=16.35, p<.001. Mean improvement: CBT = 10.09, Control = 0.6. CI between 8.35, 10.63. 9/10 point improvement 10.09 - 0.6 - Significant improvement

VR vs CBT - CBT improves more than control: t=-12.38, p<.001. Mean improvement: VR = 18.62, CBT = 10.09. CI between -9.84, -7.22. 7-10 point improvement 18.62 - 10.09

Results suggest that both CBT and VR were a significant improvement on the Control group and comparing VR and CBT showed that the VR group achieved a larger improvement

VR Therapy outperforms CBT and Control, CBT outperforms Control

**z-test results**

Evaluate whether there is a significant improvement in AQ scores across all participants, across all groups and determine whether the overall mean change in AQ scores differed significantly from 0

Does the treatment work across all groups and participants?

 - Null hypothesis - The mean change in AQ scores equals 0
 - Alternative hypothesis - The mean change in AQ scores does not equals 0
 
 
The average change in AQ scores across all participants showed a positive value (9.77) meaning that scores decreased after treatment and that participants improved

The Z score of 19.72 shows that the well above 0 and indicates that improvement is highly unlikely random and shows the strength of the evidence

P value is 0, so can reject null hypothesis

95% Confidence Interval of lower as 8.8 and upper limits at 10.74 shows that the true average improvement and the narrow confidence interval shows consistent treatment effect across the sample




#### Limitations

------------------------------------------------------------------------

# References

Maskey, M., Rodgers, J., Ingham, B., Freeston, M., Evans, G., Labus, M. and Parr, J.R. (2019). Using Virtual Reality Environments to Augment Cognitive Behavioral Therapy for Fears and Phobias in Autistic Adults. Autism in Adulthood, 1(2), pp.134–145. doi:https://doi.org/10.1089/aut.2018.0019.

De Luca, R., Leonardi, S., Portaro, S., Le Cause, M., De Domenico, C., Colucci, P.V., Pranio, F., Bramanti, P. and Calabrò, R.S. (2019). Innovative use of virtual reality in autism spectrum disorder: A case-study. Applied Neuropsychology: Child, 10(1), pp.90–100. doi:https://doi.org/10.1080/21622965.2019.1610964.

