---

title: "Comparing the Effectiveness of VR-Based and CBT Interventions for Autism Spectrum
  Disorder"
author: "Brendan Flaherty (D00285274) & Jeremiah Spillane (D00295490)"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
---

---
# Data Analytics for Immersive Environments (COMPC9047)
## Investigating the Impact of Therapeutic Interventions on Austistic Traits using the Autism Spectrum Quotient (AQ)

**Participants:** Brendan Flaherty (D00285274) & Jeremiah Spillane (D00295490)
**Module:** Data Analytics for Immersive Environments (COMPC9047)
**Date:** `r Sys.Date()`
---

---
# Table of Contents
---

---
#Abstract
##Aim and rationale
##Participants and setting
##Experiment design 
##Results gathering
##Findings/implications


# Introduction
## Topic and Context
Briefly describe the Autism Spectrum Disorder (ASD) as a neurodevelopmental spectrum condition and introduce the concept of neurodiversity.


## Rationale
Why are we doing this study? This study investigates which method yileds greater improvements in AQ

## Hypotesis
#### **H1:** After controlling for baseline AQ, there will be no group differences in post-treatment (null-hypothesis)

#### **H2:** Both CBT abd VR therapy groups will show significant reductions in AQ scores from pre-treatment to post-treatment, while the control group will not

#### **H3:** ???

# Method
#### **Participants:** 300 adults (M = 150, F = 150), aged 12-40. Random assignment to CBT (n = 100), VR (n = 100), Control (n = 100).

#### **Design:** Mixed 3 (Group: CBT, VR, Control - between subjects) x 2 (Time: Pre, Post - within-subjects) design.

#### **Materials:** Autism Spectrum Quotient (AQ) 50-item questionnaire, scored 0-50.1.

#### **Procedure:** Each participant completed AQ before and after 12 weeks of therapy.

+ CBT: Weekly sessions on social/emotional regulation.

+ VR: Immersive simulations for social interaction.

+ Control: No intervention.

Data recorded in columns P1_Pre_Score and P2_Post_Score.

# Results

### **Data Cleaning**
```{r Data Cleaning}
# Read the data frames of the .csv file
df <- read_csv("CA1_Autism_VR_Study_Data_Final.csv")
# Check the column names
colnames(df)
# Change column names to lowercase
names(df) <- tolower(names(df))
# Inspect Structure
glimpse(df)
# Check for missing values
sapply(df, function(x) sum(is.na(x)))
# Check for values outside the scope of the questionnaire - 0 - 50.1
filter(df, p1_pre_score < 0 | p1_pre_score > 50.1 | p2_post_score < 0 | p2_post_score > 50.1)
# Check for summary of all columns
summary(select(df, p1_pre_score, p2_post_score, change_score))
# Identify change_score (post-pre AQ)
df <- df %>% mutate(change_score = p2_post_score - p1_pre_score)
```


### **Descriptive Statistics**
#### **Plots**
```{r}

```

### **Inferential Statistics**
#### **Probability**
#### **Hypothesis Testing**
#### **Regresssion Analysis**
#### **Confidence Interval**

```{r}

```

### **Statistical tests**

### **Hypothesis 1 - **

### **Hypothesis 2 - **

### **Hypothesis 3 - **

### **Magnitude and direction of results**


# Discussion
#### Outline findings and relation to the hypothesis
#### Limitations
---

# References
